---
title: "Modelos lineales"
subtitle: "Clase 14: regresion"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=300,fig.width=7)
```

# Regresión

- Modelos lineales

## Altura de padres e hijos

- Empleamos los datos de Pearson & Lee (1903), los cuales contienen 1078 observaciones de alturas de pares de padres e hijos.
  - Originalmente, los datos se encuentran en pulgadas
  - Los vamos a transformar a metros, de modo de interpretar los resultados en un métrica más común a la empleada en Chile.
  - Los datos los obtenemos desde el R packege `library(UsingR)`, con el comando `UsingR::father.son`

## Preparar datos

```{r, echo=TRUE}

# -----------------------------------------------------------------------------
# child and parents height
# -----------------------------------------------------------------------------

# -----------------------------------------------
# install package
# -----------------------------------------------

# install.packages("UsingR")


# -----------------------------------------------
# load data
# -----------------------------------------------

data_alt <- UsingR::father.son

# -----------------------------------------------
# variables
# -----------------------------------------------

library(dplyr)
r4sda::variables_table(data_alt) %>%
knitr::kable()

# -----------------------------------------------
# data description
# -----------------------------------------------

# ? UsingR::UsingR::father.son

# Description:
# 
#      1078 measurements of a father's height and his son's height.
# 
# Usage:
# 
#      data(father.son)
#      
# Format:
# 
#      A data frame with 1078 observations on the following 2 variables.
# 
#      fheight Father's height in inches
# 
#      sheight Son's height in inches
# 
# Details:
# 
#      Data set used by Pearson & Lee (1903) to investigate regression. See data set
#      ‘galton’ for data set used by Galton.
# 

# Note: metric in inches (1 inch is 2.54 cms)

# -----------------------------------------------
# get descriptives
# -----------------------------------------------

data_alt %>%
r4sda::get_desc() %>%
knitr::kable(., digits = 2)

# -----------------------------------------------
# prepare data
# -----------------------------------------------

data_alt <- data_alt %>%
            mutate(parent = (fheight*2.54)/100) %>%
            mutate(child = (sheight*2.54)/100) %>%
            dplyr::glimpse()


```

## Explorar datos

```{r, echo=TRUE}

# -----------------------------------------------------------------------------
# child and parents height
# -----------------------------------------------------------------------------

# -----------------------------------------------
# install package
# -----------------------------------------------

# -----------------------------------------------
# get descriptives
# -----------------------------------------------

data_alt %>%
r4sda::get_desc() %>%
knitr::kable(., digits = 2)

# -----------------------------------------------
# correlations
# -----------------------------------------------

data_alt %>%
corrr::correlate() %>%
corrr::shave() %>%
knitr::kable(., digits = 2)


# -----------------------------------------------
# inspect with a scatter plot (plain)
# -----------------------------------------------

library(ggplot2)
data_alt %>%
ggplot(., aes(y = child, x = parent)) + 
geom_point(alpha = .3) + 
labs(x = "parents", y = "child") +
theme_linedraw()

```

## Linea de regresión sobre figuras

```{r, echo=TRUE}

# -----------------------------------------------------------------------------
# child and parents height
# -----------------------------------------------------------------------------

# -----------------------------------------------
# inspect with a scatter plot (regression line)
# -----------------------------------------------

library(ggplot2)
data_alt %>%
ggplot(., aes(y = child, x = parent)) + 
geom_point(alpha = .3) + 
geom_smooth(method = "lm", colour = "Red") +
labs(
  y = "childs' height (mts)",
  x = "parents' height (mts)",   
  ) +
theme_linedraw()

ggplot2::ggsave('regression_01.png',
     plot = last_plot(), 
     width=10, 
     height=10, 
     units="cm",
     bg = "transparent",
     dpi = 400,
     scale= 2)


# -----------------------------------------------
# inspect with a scatter plot (with correlation)
# -----------------------------------------------

library(ggplot2)
data_alt %>%
ggplot(., aes(y = child, x = parent)) + 
geom_point(alpha = .3) + 
geom_smooth(method = "lm", colour = "Red") +
labs(x = "parents", y = "child") +
ggpubr::stat_cor(method="pearson", label.y = 1.98) +
ggpubr::stat_cor(method="pearson") +
theme_linedraw()

# -----------------------------------------------
# inspect with a scatter plot (with regression equation)
# -----------------------------------------------

library(ggplot2)
data_alt %>%
ggplot(., aes(y = child, x = parent)) + 
geom_point(alpha = .3) + 
geom_smooth(method = "lm", colour = "Red") +
labs(x = "parents", y = "child") +
ggpubr::stat_cor(method="pearson", label.y = 1.98) +
ggpubr::stat_regline_equation(label.y = 1.95) +
theme_linedraw()


```

## Ajustar regressiones

```{r, echo=TRUE}

# -----------------------------------------------------------------------------
# child and parents height
# -----------------------------------------------------------------------------

# -----------------------------------------------------------------------------
# using inches
# -----------------------------------------------------------------------------

# -----------------------------------------------
# regression as in Pearson & Lee (1903)
# -----------------------------------------------


lm(sheight ~ + 1 + fheight, data = data_alt) %>%
summary()


# -----------------------------------------------
# regression output with tidy table
# -----------------------------------------------

lm(sheight ~ + 1 + fheight, data = data_alt) %>%
broom::tidy() %>%
knitr::kable(., digits = 2)


# -----------------------------------------------------------------------------
# using meters
# -----------------------------------------------------------------------------

# -----------------------------------------------
# regression model summary using mts
# -----------------------------------------------

lm(child ~ + 1 + parent, data = data_alt) %>%
summary()


# -----------------------------------------------
# regression output with tidy table
# -----------------------------------------------

lm(child ~ + 1 + parent, data = data_alt) %>%
broom::tidy() %>%
knitr::kable(., digits = 2)

# -----------------------------------------------
# r-square
# -----------------------------------------------

lm(child ~ + 1 + parent, data = data_alt) %>%
broom::glance() %>%
knitr::kable(., digits = 2)


# -----------------------------------------------------------------------------
# using inches with  lavaan
# -----------------------------------------------------------------------------

# -----------------------------------------------
# regression model with lavaan
# -----------------------------------------------

reg_model <- '
sheight ~ b0*1       # intercept
sheight ~ b1*fheight # slope
'

# -----------------------------------------------
# fit model
# -----------------------------------------------

fit_reg <- lavaan::sem(reg_model, data = data_alt, 
             mimic="Mplus", estimator = "ML")

# -----------------------------------------------
# display results
# -----------------------------------------------

lavaan::summary(fit_reg,
  standardized=TRUE, 
  fit.measures=TRUE, 
  rsquare=TRUE)

# -----------------------------------------------------------------------------
# using meters with  lavaan
# -----------------------------------------------------------------------------

# -----------------------------------------------
# regression model with lavaan
# -----------------------------------------------

reg_model <- '
child ~ b0*1      # intercept
child ~ b1*parent # slope
'

# -----------------------------------------------
# fit model
# -----------------------------------------------

fit_reg <- lavaan::sem(reg_model, data = data_alt, 
             mimic="Mplus", estimator = "ML")

# -----------------------------------------------
# display results
# -----------------------------------------------

lavaan::summary(fit_reg,
  standardized=TRUE, 
  fit.measures=TRUE, 
  rsquare=TRUE)


```


## Chequeo de supuestos

```{r, echo=TRUE}

# -----------------------------------------------------------------------------
# normaility of errors
# -----------------------------------------------------------------------------

# -----------------------------------------------
# normality assumption of errors (Shapiro Wilk)
# -----------------------------------------------

lm(child ~ + 1 + parent, data = data_alt) %>%
residuals() %>%
shapiro.test()

# -----------------------------------------------
# normality assumption of errors (Kolmogorov Smirnoff)
# -----------------------------------------------

residual_values <- residuals(lm(child ~ + 1 + parent, data = data_alt))
fitted_values   <- predict(lm(child ~ + 1 + parent, data = data_alt))

data_hom <- data_alt %>%
            mutate(fit = fitted_values) %>%
            mutate(errors = residual_values)


ks.test(data_hom$errors, 
  "pnorm", 
  mean=mean(data_hom$errors), 
  sd=sd(data_hom$errors)
  )


# -----------------------------------------------
# normality plot of errors
# -----------------------------------------------

error_mean <- mean(data_hom$errors, na.rm = TRUE)
error_sd <- sd(data_hom$errors, na.rm = TRUE)
error_n  <- nrow(data_hom$errors)
binwidth <- 0.01

data_hom %>%
ggplot(., aes(x = errors, mean = error_mean, sd = error_sd, binwidth = binwidth, n = error_n)) +
    theme_linedraw() +
    geom_histogram(
        aes(y =..density..),
        binwidth = binwidth, 
        colour = "white", 
        fill = "grey60", 
        size = 0.1) +
stat_function(fun = dnorm, args = list(mean = error_mean, sd = error_sd), geom = "line")


# -----------------------------------------------
# normality of residuals, qqplot
# -----------------------------------------------


residual_values <- residuals(lm(child ~ + 1 + parent, data = data_alt))

data_residual <- data_alt %>%
                 mutate(errors = residual_values)


ggplot(data_residual, aes(sample = errors)) +
  stat_qq(alpha = .5) +
  labs(y = "residuals") +
theme_linedraw()


# -----------------------------------------------
# homocesdasticiy plot
# -----------------------------------------------

residual_values <- residuals(lm(child ~ + 1 + parent, data = data_alt))
fitted_values   <- predict(lm(child ~ + 1 + parent, data = data_alt))

data_hom <- data_alt %>%
            mutate(fit = fitted_values) %>%
            mutate(errors = residual_values)


data_hom %>%
ggplot(., aes(y = errors, x = fit)) + 
geom_point(alpha = .3, col = 'grey20') + 
# geom_segment(aes(xend = parent, yend = fit), alpha = .4, col = 'red') + 
# geom_point(aes(y = fit), col = 'grey25', alpha = .5) + 
xlab('Fitted Values') +
ylab('Residuals') + 
ggtitle('Residual Dispersion') +
geom_hline(yintercept = 0, linetype=2)+
theme_linedraw()

# -----------------------------------------------
# homocesdasticiy Breusch-Pagan Test
# -----------------------------------------------

lm(child ~ + 1 + parent, data = data_alt) %>%
lmtest::bptest()

```

# Anexos

## Plots empleados en clases

```{r, echo=TRUE}

# -----------------------------------------------------------------------------
# child and parents height
# -----------------------------------------------------------------------------

# -----------------------------------------------
# install package
# -----------------------------------------------

# install.packages("UsingR")

# -----------------------------------------------
# load data
# -----------------------------------------------

data_alt <- UsingR::father.son

# -----------------------------------------------
# prepare data
# -----------------------------------------------

data_alt <- data_alt %>%
            mutate(parent = (fheight*2.54)/100) %>%
            mutate(child = (sheight*2.54)/100) %>%
            dplyr::glimpse()


# -----------------------------------------------
# inspect with a scatter plot (regression line)
# -----------------------------------------------

library(ggplot2)
data_alt %>%
ggplot(., aes(y = sheight, x = fheight)) + 
geom_point(alpha = .3) + 
geom_smooth(method = "lm", colour = "Red") +
labs(x = "parents", y = "child") +
ggpubr::stat_regline_equation(label.y = 80.00) +
theme_linedraw()

ggplot2::ggsave('regression_01_original.png',
     plot = last_plot(), 
     width=10, 
     height=10, 
     units="cm",
     bg = "transparent",
     dpi = 400,
     scale= 2)


# -----------------------------------------------
# regression as in Pearson & Lee (1903)
# -----------------------------------------------

lm(sheight ~ + 1 + fheight, data = data_alt) %>%
broom::tidy() %>%
knitr::kable(., digits = 2)

# -----------------------------------------------
# regression line in mts
# -----------------------------------------------

library(ggplot2)
data_alt %>%
ggplot(., aes(y = child, x = parent)) + 
geom_point(alpha = .3) + 
geom_smooth(method = "lm", colour = "Red") +
labs(
  y = "childs' height (mts)",
  x = "parents' height (mts)",   
  ) +
theme_linedraw()

ggplot2::ggsave('regression_01_mts.png',
     plot = last_plot(), 
     width=10, 
     height=10, 
     units="cm",
     bg = "transparent",
     dpi = 400,
     scale= 2)


# -----------------------------------------------
# plot with full axis
# -----------------------------------------------

library(ggplot2)
data_alt %>%
ggplot(., aes(y = sheight, x = fheight)) + 
geom_point(alpha = .1) + 
geom_smooth(method = "lm", colour = "Red") +
labs(x = "parents", y = "child") +
ggpubr::stat_regline_equation(label.y = 80.00) +
xlim(0,80) +
ylim(0,80) +
theme_linedraw()

ggplot2::ggsave('regression_01_full_axis.png',
     plot = last_plot(), 
     width=10, 
     height=10, 
     units="cm",
     bg = "transparent",
     dpi = 400,
     scale= 2)

lm(sheight ~ + 1 + fheight, data = data_alt) %>%
broom::tidy() %>%
knitr::kable(., digits = 2)



# -----------------------------------------------
# plot with full axis
# -----------------------------------------------

fheight_mean <- mean(data_alt$fheight, na.rm = TRUE)

library(ggplot2)
data_alt %>%
mutate(fheight_centered = fheight - fheight_mean) %>%
ggplot(., aes(y = sheight, x = fheight_centered)) + 
geom_point(alpha = .1) + 
geom_smooth(method = "lm", colour = "Red") +
labs(x = "parents", y = "child") +
ylim(60,80) +
xlim(-10,10) +
ggpubr::stat_regline_equation(label.y = 80.00) +
theme_linedraw()

ggplot2::ggsave('regression_01_centered.png',
     plot = last_plot(), 
     width=10, 
     height=10, 
     units="cm",
     bg = "transparent",
     dpi = 400,
     scale= 2)

data_alt %>%
mutate(fheight_centered = fheight - fheight_mean) %>%
lm(sheight ~ + 1 + fheight_centered, data = .) %>%
broom::tidy() %>%
knitr::kable(., digits = 2)


# -----------------------------------------------
# residuals in comparison to the model
# -----------------------------------------------


fitted_values <- predict(lm(sheight ~ + 1 + fheight, data = data_alt))

data_fitted <- data_alt %>%
               mutate(fit = fitted_values)

data_fitted %>%
ggplot(., aes(y = sheight, x = fheight)) + 
geom_point(alpha = .3, col = 'grey20') + 
geom_segment(aes(xend = fheight, yend = fit), alpha = .4, col = 'red') + 
geom_point(aes(y = fit), col = 'grey25', alpha = .5) + 
xlab('Height of Father') +
ylab('Height of Sons') + 
ggtitle('Residual Distance from the Model') +
theme_linedraw()


ggplot2::ggsave('regression_01_residuals.png',
     plot = last_plot(), 
     width=10, 
     height=10, 
     units="cm",
     bg = "transparent",
     dpi = 400,
     scale= 2)


# -----------------------------------------------
# normality assumption of errors (Kolmogorov Smirnoff)
# -----------------------------------------------

residual_values <- residuals(lm(child ~ + 1 + parent, data = data_alt))
fitted_values   <- predict(lm(child ~ + 1 + parent, data = data_alt))

data_hom <- data_alt %>%
            mutate(fit = fitted_values) %>%
            mutate(errors = residual_values)


ks.test(data_hom$errors, 
  "pnorm", 
  mean=mean(data_hom$errors), 
  sd=sd(data_hom$errors)
  )


# -----------------------------------------------
# normality plot of errors
# -----------------------------------------------

residual_values <- residuals(lm(child ~ + 1 + parent, data = data_alt))
fitted_values   <- predict(lm(child ~ + 1 + parent, data = data_alt))

data_hom <- data_alt %>%
            mutate(fit = fitted_values) %>%
            mutate(errors = residual_values)

error_mean <- mean(data_hom$errors, na.rm = TRUE)
error_sd <- sd(data_hom$errors, na.rm = TRUE)
error_n  <- nrow(data_hom$errors)
binwidth <- 0.01

data_hom %>%
ggplot(., aes(x = errors, mean = error_mean, sd = error_sd, binwidth = binwidth, n = error_n)) +
    theme_linedraw() +
    geom_histogram(
        aes(y =..density..),
        binwidth = binwidth, 
        colour = "white", 
        fill = "grey60", 
        size = 0.1) +
stat_function(fun = dnorm, args = list(mean = error_mean, sd = error_sd), geom = "line")



ggplot2::ggsave('regression_04_normality.png',
     plot = last_plot(), 
     width=10, 
     height=10, 
     units="cm",
     bg = "transparent",
     dpi = 400,
     scale= 2)


# -----------------------------------------------
# homocesdasticiy plot
# -----------------------------------------------

residual_values <- residuals(lm(child ~ + 1 + parent, data = data_alt))
fitted_values   <- predict(lm(child ~ + 1 + parent, data = data_alt))

data_hom <- data_alt %>%
            mutate(fit = fitted_values) %>%
            mutate(errors = residual_values)


data_hom %>%
ggplot(., aes(y = errors, x = fit)) + 
geom_point(alpha = .3, col = 'grey20') + 
# geom_segment(aes(xend = parent, yend = fit), alpha = .4, col = 'red') + 
# geom_point(aes(y = fit), col = 'grey25', alpha = .5) + 
xlab('Fitted Values') +
ylab('Residuals') + 
ggtitle('Residual Dispersion') +
geom_hline(yintercept = 0, linetype=2)+
theme_linedraw()



ggplot2::ggsave('regression_05_homocedasticity.png',
     plot = last_plot(), 
     width=10, 
     height=10, 
     units="cm",
     bg = "transparent",
     dpi = 400,
     scale= 2)


```



